---
title: "Project 3 - Weather Application Continued"
format: 
  html:
    self-contained: true
    toc: true
    toc-depth: 4
    toc-location: right-body
author:
   - name: Andrew Moles
     affiliation: Learning Developer, Digital Skills Lab
date: today    

---

## Learning objectives:
- Create vectors
- Use calculations on vectors
- Extract elements from vectors
- Using boolean indexing on vectors
- Print out hourly weather reports using iteration


## Outcomes

We will continue to write a program that tells us various metrics on what the weather is doing. Your code will automate the process of taking data and making it into a presentable format. 

We will be aiming for two outcomes in this project.

### Outcome 1 - Make a hourly weather report:

You should end up with a output that iterates through each element of your temperatures and times data. Each line should be an hour with the temperature in Celsius and Fahrenheit (in brackets). 

It should also print out what the date at the top of your report. 

```{r}
#| echo: false

temps <- c(20.8, 19.6, 18.7, 17.8, 17.7, 19.1, 21.4, 22.8, 24.1, 26.5, 27.8, 
           29.4, 30.7, 31.8, 32.4, 32.0, 31.8, 31.5, 31.2, 29.9, 28.3, 27.1, 26.2)
dates <- c("30/06/2025 01:00:00", "30/06/2025 02:00:00", "30/06/2025 03:00:00", 
           "30/06/2025 04:00:00", "30/06/2025 05:00:00", "30/06/2025 06:00:00",
           "30/06/2025 07:00:00", "30/06/2025 08:00:00", "30/06/2025 09:00:00",
           "30/06/2025 10:00:00", "30/06/2025 11:00:00", "30/06/2025 12:00:00",
           "30/06/2025 13:00:00", "30/06/2025 14:00:00", "30/06/2025 15:00:00",
           "30/06/2025 16:00:00", "30/06/2025 17:00:00", "30/06/2025 18:00:00",
           "30/06/2025 19:00:00", "30/06/2025 20:00:00", "30/06/2025 21:00:00",
           "30/06/2025 22:00:00", "30/06/2025 23:00:00")

#temperature_convert <- function(temp_data, unit) {
#  if (unit == "F") {
#    temp_data <- (temp_data * 9/5) + 32
#  } else if (unit == "C") {
#    temp_data <- (temp_data - 32) * 5/9
#  } else {
#    print("Please provide unit you want to convert to of °C or °F")
#  }
#  return(temp_data)
#}

dmy <- substr(dates[1], 1, 10)
temps_f <- (temps * 9/5) + 32
hour <- substr(dates, 12, 13)

# classic print-paste method
print(paste0("Hourly temperatures on: ", substr(dates[1], 1, 10)))
for (i in 1:length(temps)) {
  print(
    paste0("Temperature at ", hour[i], ": ",  temps[i], "°C (", temps_f[i], "°F)")
    )
}

# using cat instead of print-paste (personal fav)
#for (i in 1:length(temps)) {
#  cat("Temperature at ", hour[i], ": ",  temps[i], "°C (", temps_f[i], "°F)\n",
#      sep = "")
#}

# the same but using the recommended seq_along method
#cat("Hourly temperatures on: ", substr(dates[1], 1, 10), sep = "")
#for (i in seq_along(temps)) {
#    cat("Temperature at ", hour[i], ": ",  temps[i], "°C (", temps_f[i], "°F)\n",
#      sep = "")
#}

```


### Outcome 2 - Finding hours with high temperatures

We can filter our temperature vector, which we can add names to, which tells us which hours had temperatures over 30°C. 

You should end up with an output, like below, that has the hour information and the temperature information in one vector. 

```{r}
#| echo: false

names(temps) <- hour
temps[temps > 30]
```


### The data

Using the data below we will be making a weather reporting tool for a hourly report.

**Hourly temperatures:**

`20.8, 19.6, 18.7, 17.8, 17.7, 19.1, 21.4, 22.8, 24.1, 26.5, 27.8, 29.4, 30.7, 31.8, 32.4, 32.0, 31.8, 31.5, 31.2, 29.9, 28.3, 27.1, 26.2`

**Date and times:**

`"30/06/2025 01:00:00", "30/06/2025 02:00:00", "30/06/2025 03:00:00", "30/06/2025 04:00:00", "30/06/2025 05:00:00", "30/06/2025 06:00:00", "30/06/2025 07:00:00", "30/06/2025 08:00:00", "30/06/2025 09:00:00", "30/06/2025 10:00:00", "30/06/2025 11:00:00", "30/06/2025 12:00:00","30/06/2025 13:00:00", "30/06/2025 14:00:00", "30/06/2025 15:00:00", "30/06/2025 16:00:00", "30/06/2025 17:00:00", "30/06/2025 18:00:00","30/06/2025 19:00:00", "30/06/2025 20:00:00", "30/06/2025 21:00:00", "30/06/2025 22:00:00", "30/06/2025 23:00:00"`


## Steps to help you get to the outcome

### Part 1 - the setup
Open an R script file and save it.

### Part 2 - using vectors
Using the data shown above, store the hourly temperatures and date and times data into R as vectors.

### Part 3 - string manipulation
Extract part of the string from the date and time data to give you just the date element. Store the result as a variable, keeping just the first element; you should get the result below. 

`r dmy`

We want to do this programmatically, so no copy and pasting from the data, also known as "hardcoding". This will help us to automate our code if we want to use different data. 

### Part 4 - more string manipulation
Extract part of the string from the date and time data to give you just the hour elements. Store this as a vector. 

### Part 5 - convert temperatures
Convert the hourly temperatures data to Fahrenheit and store the result as a new vector.

### Part 6 - start making the text output
Our ultimate task is to iterate through our temperature and date data to print out a message telling us the temperature for each hour. To get there it is sometimes helpful to breakdown the process. 

Try first to a make an output for just the first hour, like the result below, using vector indexing to get the data you need. Then use string concatenation to bring all the information together to create a text output. 

`r paste("Temperature at ", hour[1], ": ",  temps[1], "°C (", temps_f[1], "°F)\n", sep = "")`

### Task 7 - finish the text output
Now using the code you just wrote, iterate your code so you get the hourly weather report with both Celsius and Fahrenheit.

### Task 8 - naming a vector

Using vector naming, add the hours vector as a name to the temperature vector.

### Part 9 - filtering the temperature vector

Using boolean indexing, filter the temperature vector so we are only left with temperatures over 30°C.

### Part 10 - test your code with new data
A good way to test out your code is to try it on different data. 

Below we have the temperature data for a few days after the data we have been using. Using the steps you did previously, write a weather report using this new data. 

**Hourly temperatures:**

`16.5, 15.5, 14.3, 13.7, 12.4, 12, 12.6, 14.1, 16, 18.2, 19.7, 21.1, 21.8, 22.6, 23.4, 24.3, 23.3, 23.6, 23.1, 23.4, 23.4, 21.6, 20.7, 19.5`

**Date and times:**

`"03/07/2025 00:00:00", "03/07/2025 01:00:00", "03/07/2025 02:00:00", "03/07/2025 03:00:00", "03/07/2025 04:00:00", "03/07/2025 05:00:00", "03/07/2025 06:00:00", "03/07/2025 07:00:00", "03/07/2025 08:00:00", "03/07/2025 09:00:00", "03/07/2025 10:00:00", "03/07/2025 11:00:00", "03/07/2025 12:00:00", "03/07/2025 13:00:00", "03/07/2025 14:00:00", "03/07/2025 15:00:00", "03/07/2025 16:00:00", "03/07/2025 17:00:00", "03/07/2025 18:00:00", "03/07/2025 19:00:00", "03/07/2025 20:00:00", "03/07/2025 21:00:00", "03/07/2025 22:00:00", "03/07/2025 23:00:00"`

### Final task - fill out the survey!


We are always looking to improve and iterate our workshops. [Follow the link to give your feedback](https://lse.eu.qualtrics.com/jfe/form/SV_0vJs5yGpF1F359I?coursename=R Projects 3: Building a Weather Application Continued&topic=R&prog=DS&version=25-26){target="_blank"}.


 <!--
### Optional outcome

Can you also get the result to be a vector?

```{r}
#| echo: false
# vector version
paste0("Hourly temperatures on: ", substr(dates[1], 1, 10))
paste0("Temperature at ", substr(dates, 12, 13), ": ",  temps, "°C (", temps_f, "°F)")

```

--->


